---
title: "03: phyloseq"
author: "Sarah Teichman"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Phyloseq

`phyloseq` is an R package that is used to store, analyze, and plot microbiome abundance data. We will start by importing it from the Bioconductor package repository. 

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

if (!requireNamespace("phyloseq", quietly = TRUE)) {
     BiocManager::install("phyloseq")
}

# load in the phyloseq package
library(phyloseq)

# load in tidyverse
library(tidyverse)
```

If you have any issues installing `phyloseq` on your local computer, we recommend switching over to your R Studio server account where `phyloseq` is already installed.

## Load in phyloseq object 

We'll be using the same data as in our `covariates` and `otus` data frames that we've working with, but now in a different form. To start, we will load in our `phyloseq` object. 

```{r}
phyloseq_obj <- readRDS("data/phyloseq_obj.rds")
```

## Using phyloseq objects

Great, we have a `phyloseq` object! But what is it?

A `phyloseq` object is made up of three tables. 

1. Sample data table
2. OTU count table
3. Taxonomy table

This lets us link all of our information about our data (covariates for each sample, otus counts, and taxonomic information about OTUs) into a single object. That's pretty cool! 

Let's start by looking at the individual components of the object. 

```{r}
# sample data
head(sample_data(phyloseq_obj))

# otu count data 
head(otu_table(phyloseq_obj))

# taxonomy data 
head(tax_table(phyloseq_obj))
```

Once we have our `phyloseq` object, we can do a lot with it. Let's start out by learning the basics of it (which we probably already know from our explorations of the data). 

```{r}
# check out our sample data 
nsamples(phyloseq_obj)
sample_names(phyloseq_obj)
sample_variables(phyloseq_obj)

# check out our otu count data 
sample_sums(phyloseq_obj)
ntaxa(phyloseq_obj)
head(taxa_sums(phyloseq_obj), 20)

# check out our taxonomy table
rank_names(phyloseq_obj)
```

We can also change the level of taxonomic classification we are working with. What if we only cared about differences up to the order level? 

```{r}
# use tax_glom to aggregate, where the second argument is the taxonomic level to aggregate to 
phylo_order <- tax_glom(phyloseq_obj, "order")
ntaxa(phylo_order)
head(tax_table(phylo_order))
# note that now the family and genus levels are all NA value because we have 
# aggregated to the phylum level 
# NA is a data type in R that refers to missing data or data that doesn't exist
```

What if we only want to look at taxa from the phylum *Proteobacteria*? We can also use `phyloseq` to subset taxa. 

```{r}
# use subset_taxa to subset, using the second argument to specify which taxa you want at which taxonomic rank
phylo_proteo <- subset_taxa(phyloseq_obj, phylum == "Proteobacteria")
ntaxa(phylo_proteo)
head(tax_table(phylo_proteo))
```

## Plotting phyloseq objects

One very useful aspect of `phyloseq` is its analysis and plotting capacity. Although you could make any plot you want with the two data sets and some time spent with `dplyr` and `ggplot` (which you'll learn in the next tutorials), `phyloseq` makes this process more convenient for you. 

Imagine we want to look into alpha diversity for our data. 

```{r}
# estimate several types of alpha diversity 
estimate_richness(phyloseq_obj)
# plot Shannon and Simpson diversity measures by location and season
plot_richness(phyloseq_obj, x = "Location", color = "Season", measures = c("Shannon", "Simpson")) 
```

Now imagine we want to make an ordination plot where we compare communities based on Bray-Curtis dissimilarities. 

```{r}
# calculate bray-curtis dissimilarities between each sample 
# then perform dimension reduction with nMDS (non-metric Multi Dimensional Scaling)
phylo_bray <- ordinate(phyloseq_obj, "NMDS", "bray")
# color our plot by season 
plot_ordination(phyloseq_obj, phylo_bray, type="samples", color="Season") 
```

Exercise: What is an insight about our data that we can draw from one of these two plots?

Exercise: Using the `phyloseq_obj` object, subset to only taxa that belong to *Firmicutes*. Build an ordination plot based on the subsetted `phyloseq` object you've created using Bray-Curtis dissimilarities and nMDS (as in the above section) and color the points by the location where they were collected.

*Hint*: subset to a specific taxa with `subset_taxa()`. 

*Hint*: locations are saved as the "Location" variable in the `covariates` dataset.

```{r}

```

## Getting data out of phyloseq

Finally, there are some settings where you might rather work with individual data tables than with your phyloseq object. Pretend for a moment that you started with your `phyloseq_obj` object and have no access to the `covariates` or `abundances` data frames. `phyloseq` has handy functions for us to extract each of its three tables. 

```{r}
# extract sample data table
samp_tab <- sample_data(phyloseq_obj)
head(samp_tab)

# extract otu table
otu_tab <- otu_table(phyloseq_obj)
head(otu_tab)

# extract taxonomy table
taxon_tab <- tax_table(phyloseq_obj)
head(taxon_tab)
```

Now you have each of these tables as their own R objects and can manipulate and analyze them however you would like!

## Bonus - creating your own phyloseq object 

The following code creates the object `phyloseq_obj` from our data tables `covariates` and `otus`. Feel free to follow along to see how we did this (but also feel free to move on to the next tutorial)! 

```{r}
covariates <- read.csv("data/covariates.txt", sep = "\t")
otus <- read.csv("data/otus.txt", sep = "\t", row.names = 1, header = T)
# our covariates data frame is our sample data table. 
# let's move the "SampleName" column to be row names for our data frame
row.names(covariates) <- covariates$SampleName
covariates <- covariates %>% select(-SampleName)

# our abundance table doesn't need to have all of the taxonomic information. 
# let's update the row names by numbering our OTUs. 
# first we'll save the current rownames, because we need that taxonomy information later
taxonomy <- row.names(otus)
# replace row names with "taxa1" up through "taxa387" with the "paste0()" function
row.names(otus) <- paste0("taxa", 1:nrow(otus))

# now we can make a taxonomy table
# we would like to separate each taxonomic identification into separate columns for each hierarchical level 
# make a data frame for taxonomy 
tax_df <- data.frame(tax = taxonomy)
# separate the single column tax by creating a new column each time a ";" is seen
# name the newly created columns based on their rank 
tax_df_sep <- separate_wider_delim(data = tax_df, cols = tax, delim = ";",
                            names = c("kingdom", "phylum", "class",
                                      "order", "family", "genus"))
# turn the data frame into a matrix 
# (phyloseq prefers a matrix of taxonomy to a data frame)
# (I don't know why)
tax_mat <- as.matrix(tax_df_sep)
# make sure the row names of the taxonomy matrix match the row names of the OTU table 
row.names(tax_mat) <- row.names(otus)
# now we can finally make a phyloseq object! 
phyloseq_obj <- phyloseq(otu_table(otus, taxa_are_rows = TRUE), 
                      tax_table(tax_mat), 
                      sample_data(covariates))
phyloseq_obj
```

## References 

```{r}
citation("phyloseq")
```

Some materials adapted from this [tutorial](https://www.nicholas-ollberding.com/post/introduction-to-phyloseq/) by Nicholas Ollberding.