---
title: "An Introduction to R"
author: "Sarah Teichman"
date: ''
output:
  ioslides_presentation: default
  beamer_presentation: default
  slidy_presentation: default
subtitle: The basics
---

## What is R

R is a programming language commonly used for data analysis and statistics.

- reproducible
- free
- open-source
- large community of users and developers 

Download [here](https://cran.rstudio.com/).

## What is RStudio

RStudio is an Integrated Development Environment (IDE) for R. 

- write code
- run code 
- navigate files
- visualize plots 
- open help files

Download [here](https://posit.co/download/rstudio-desktop/).

## How to access RStudio

- locally, if you have downloaded R and RStudio on your computer
  - uses your own computer and with access to your files
  - has the computational resources and limitations of your computer
- remotely, via RStudio server 
  - access via a web browser 
  - often available through your institution 
  - may have computational advantages
  - useful in a course setting like this one! 
  
## How to access RStudio server

- find your personal R Studio server link from [this spreadsheet](https://github.com/mblstamps/stamps2024/wiki/accessing-our-cloud-computers)
- username: stamps
- password: stamps2024

## RStudio organization 

RStudio has a four pane layout. 

- console (run single lines of code)
- editor (open and write scripts)
- environment etc. (see what objects exist in work space)
- files etc. (navigate files, view plots, open help files)

## Console

Use the **console** to run individual lines of code.

```{r}
5 + 384
```

```{r}
x <- 10 # set variable with <- operator :) 
y = 6 # set variable with = operator :( 
x + y
```

## Editor

Use the **editor** for opening and writing scripts.

- for a workflow to be reproducible, all code should be written in a script (not in the console)
- in R you are working in a folder on your computer 
  - `getwd()` to see (get) your working directory
  - `setwd()` to change (set) your working directory
- run code with `Run` button (and options) or `Ctrl`/`Command` + `Enter` for a single line 
  
## Environment and History

- each object saved in your working space will be in the **environment**
- **history** saves most recent lines of code
- extension: you can add a **Git** plug-in to this pane for version control through GitHub
  - [here](https://happygitwithr.com/index.html) is a great resource for R and Git!

## Files/Plots/Help 

- use files to navigate files on your computer
- use plots to display visualizations
- use help to access help files
  - type `?` to pull up a file, for example `?sum`
  - for more extensive questions, Google is also useful! 
  
## Packages

- base functions (Base R) are automatically installed with R
  - includes mathematical operations, data manipulation, plotting, etc. 
- a package is a way to store files with code, documentation, and data, and let users download and use those files
- the `tidyverse` is a suite of common data manipulation and visualization packages
  - includes `dplyr`, `ggplot`, among others

## R Markdown 

- in R scripts (.R), each line is evaluated unless it is a comment 
  - `# this is a comment` 
- in R Markdown files (.Rmd), you can combine code, output, and text
  - code in "chunks", anything within chunk is evaluated 
  - anything outside of chunk is output as text
- when compiled or "knit" .Rmd files turn into HTML, PDF, slides, webpages, etc.

## What to do when you're stuck 

- no one (that I know) writes code perfectly the first time! 
- when I get an error message, I'll often start by Googling the error
- you could also ask ChatGPT for help with your code
  - "write me an R script to use ggplot to make a histogram of the variable 'count' in the data frame 'data'"
- here at STAMPS, chat with your neighbors and get help from the instructional team! 

## This session

- download the "intro_R.zip" file from ... and upload it to your remote RStudio instance
- change your knit directory to be your "Current Directory" following the instructions below
  - in your RStudio instance, click "Tools" from the top bar, then "Global Options"
  - click on "R Markdown" in the left pane
  - find the "Evaluate chunks in directory:" and select "Current Directory" from the drop-down
  - click "Ok" to save the changes and exit


## This session
- work through the file "intro-to-R.Rmd" and the accompanying tutorials
  - this is self-paced, start where you think will be the most helpful to you
- put your stickie notes up with questions!
